---
title: "Analyzing Crime Trends in Toronto: Insights from the Police Annual Statistical Report"
# subtitle:
author: 
  - Yizhe Chen
thanks: "Code and data are available at: https://github.com/YizheChenUT/Analyzing-Crime-Trends-in-Toronto-Insights-from-the-Police-Annual-Statistical-Report.git"
date: today
date-format: long
abstract: "This paper analyzes crime trends in Toronto using the Police Annual Statistical Report from 2014 to 2023. Data from Open Data Toronto is used to evaluate crime distributions across different categories and regions. Analysis reveals that some areas report the highest number of property-related crimes like auto theft. Suburban areas also illustrate rising trends in specific crime subtypes. These findings can help optimize policing strategies and public safety resource allocations."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
library(gt)
```


# Introduction
Crime and public safety are critical concerns for urban environments like Toronto. With a growing population and increasing urbanization, it is vital to understand the nature and distribution of crimes. Crime statistics offer insights into the effectiveness of policing efforts, trends in specific types of crimes, and the socio-economic factors that may influence these trends. Despite the availability of crime data through public portals like Open Data Toronto, gaps remain in how this data is analyzed and utilized to address public safety concerns [@OpenDataDataset]. Additionally, the dynamic nature of urban crime necessitates an ongoing, detailed analysis to ensure that resource allocations remain effective and that public safety strategies are continually optimized.

This paper focuses on analyzing crime trends in Toronto from 2014 to 2023 using the Toronto Police Annual Statistical Report. Specifically, it examines two broad categories of crime: Crimes Against Property (such as auto theft and break-ins) and Crimes Against the Person (such as assault). The primary objective of this analysis is to evaluate how these crimes are distributed across Toronto's police divisions and how effectively law enforcement is addressing these incidents, measured by clearance rates (the percentage of resolved cases).

A clear gap exists in understanding how different divisions are affected by varying types of crimes and how effectively these crimes are being addressed. This analysis aims to fill that gap by offering a division-level examination of crime trends, using detailed statistical tools to identify key areas where law enforcement resources may be optimized. The paper also explores the relationship between crime types and clearance rates, highlighting divisions where law enforcement may struggle to address specific types of crimes, such as property offenses, which tend to have lower resolution rates.

In conducting this analysis, data from the Open Data Toronto portal was cleaned, standardized, and analyzed using R, revealing significant differences in crime trends between Toronto's police divisions [@citeR]. The findings indicate that certain divisions, particularly those in urban centers, report higher incidences of property crimes, while more suburban divisions face a higher prevalence of violent crimes. Moreover, clearance rates for property crimes, such as auto theft, remain low, indicating a need for enhanced crime prevention and resolution strategies in these areas. These insights are critical for shaping future public safety policies and allocating law enforcement resources more effectively.

The remainder of this paper is structured as follows: @sec-data presents the data and the steps taken to clean and transform it for analysis. @sec-result discusses the results of the study, focusing on crime distribution trends by year, types of crimes by division, and clearance rates for different crime types. @sec-discussion interprets the findings in the broader context of urban crime trends and offers insights for future research. Finally, @sec-conclusion concludes the paper by summarizing key takeaways and recommending strategies for crime prevention and law enforcement improvement in Toronto.

# Data {#sec-data}
The data used in this analysis was obtained from the Open Data Toronto portal [@OpenDataDataset]. Data was cleaned and analyzed using the language R [@citeR]. The data set includes reported incidents categorized by type, division, and year across Toronto, covering the period from 2014 to 2023. The data collection is a key resource for tracking public safety, and its availability promotes transparency in policing efforts. However, the dataset also raises ethical concerns, particularly around data privacy. Although anonymized, the geospatial breakdown could still enable the identification of patterns in specific communities, potentially stigmatizing certain areas.

## Reported Crimes Dataset
The dataset includes multiple fields related to crime, such as:  
**Crime Category**: Broad crime categories like "Crimes Against Property" and "Crimes Against the Person".  
**Crime Subtype**: Specific types of crimes (e.g., "Auto Theft", "Break & Enter").  
**Division**: The Toronto Police divisions where crimes occurred.  
**Incidents Reported**: The total number of reported incidents.  
**Incidents Cleared**: The number of resolved cases.

## Data Cleaning and Transformation
To prepare the dataset for analysis, I performed a thorough cleaning process using the janitor package [@janitor], which removed duplicates and corrected entry errors. Given the size and complexity of the dataset, we also used dplyr [@tidyverse] to filter, arrange, and group data based on specific crime subtypes and divisions. @fig-Crimes_by_Division illustrates the breakdown of crimes across different divisions from 2014 to 2023.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

crime_data <- read_csv(here::here("data/analysis_data/analysis_data.csv")) # Load the cleaned analysis data

## Filter and organize data for the plot
crime_by_division <- crime_data %>%
  group_by(division_name, year) %>%
  summarize(total_crimes = sum(incidents_reported, na.rm = TRUE))
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-Crimes_by_Division
#| fig-cap: Total Crimes across Different Divisions

## Plot crime trends by division
crime_by_division %>%
  ggplot(aes(x = year, y = total_crimes, color = division_name)) +
  geom_line(size = 1) +
  labs(
    title = "Crime Incidents by Division (2014-2023)",
    x = "Year",
    y = "Total Incidents",
    color = "Division"
  ) +
  theme_minimal() # Minimal theme for clean visuals
```

## Broader Context and Ethical Considerations
The crime data used in this report exists within a broader societal context, involving both ethical and statistical issues. Ethically, the publication of crime statistics is a double-edged sword. While transparency in crime reporting helps the public and policymakers make informed decisions, there are concerns that reporting on crime rates by division may reinforce stereotypes or unjustly stigmatize certain neighborhoods.

From a statistical view, ensuring the accuracy of reported crimes is vital for maintaining public trust. Misreporting or underreporting of crimes, either by victims or authorities, can skew the perception of public safety and crime trends. The dataset itself does not account for unreported crimes, which is a limitation that should be considered when interpreting the results. Further, as crime rates fluctuate over time, the consistency of data collection methodologies is essential for making accurate year-over-year comparisons.

## Graphical Representation and Tables
To have a better understanding of crime trends, @fig-Crimes_by_Division and @fig-Total_Crimes_by_year below present detailed graphical representations of the actual data, generated using ggplot2 [@ggplot2]. The graphs highlight crime distributions across divisions and time periods, avoiding the use of summary statistics to give a more detailed view of crime trends. @tbl-Total_Crimes_by_year_div, created using the gt package (@gt), provides a tabular summary of crime data by year and division.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

## Group data by year for total crime counts
crime_by_year <- crime_data %>%
  group_by(year) %>%
  summarize(total_crimes = sum(incidents_reported, na.rm = TRUE))
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-Total_Crimes_by_year
#| fig-cap: Total Crimes by Year

## Plot crime trends by year
crime_by_year %>%
  ggplot(aes(x = year, y = total_crimes)) +
  geom_line(size = 1, color = "blue") +
  labs(
    title = "Crimes Reported by Year (2014-2023)",
    x = "Year",
    y = "Total Crimes Reported"
  ) +
  theme_minimal() # Minimal theme for clean visuals
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

## Prepare data for table
crime_summary_table <- crime_data %>%
  group_by(year, division_name) %>%
  summarize(total_crimes = sum(incidents_reported, na.rm = TRUE))
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-Total_Crimes_by_year_div
#| tbl-cap: Total Crimes by Year and Division

## Create a table using gt
crime_summary_table %>%
  gt() %>%
  tab_header(
    title = "Crime Incidents by Year and Division (2014-2023)"
  ) %>%
  cols_label(
    year = "Year",
    division_name = "Police Division",
    total_crimes = "Total Crimes Reported"
  )
```

# Results {#sec-result}
## Crime Distribution by Year
The dataset reveals fluctuating crime rates between 2014 and 2023. Property crimes such as auto theft have shown consistent increases, while personal crimes, such as sexual violations, remain relatively stable.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv")) # Load the cleaned analysis data
crime_trends_plot <- analysis_data %>%
  group_by(year, crime_category) %>%
  summarise(total_incidents = sum(incidents_reported, na.rm = TRUE)) # Summing incidents per year and crime category
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-Crimes_by_Year
#| fig-cap: Crimes Reported by Year

## Create a line chart showing crimes reported by year and crime category
## Citation: https://allisonhorst.github.io/palmerpenguins/
## Citation: https://ggplot2.tidyverse.org
crime_trends_plot %>%
  ggplot(aes(x = year, y = total_incidents, color = crime_category, group = crime_category)) +
  geom_line(size = 1) + # Line plot for crime trends
  geom_point(size = 2) + # Adding points to lines
  labs(
    title = "Crimes Reported by Year in Toronto (2014-2023)",
    x = "Year",
    y = "Total Incidents Reported",
    color = "Crime Category"
  ) +
  theme_minimal() # Minimal theme for clean visuals
```

According to @fig-Crimes_by_Year, There was a sharp increase in crimes against property in 2023, reaching the highest number of incidents reported during the 2014-2023 period, with over 100,000 incidents. Crimes against the person maintained a relatively stable trend from 2014 to 2023, consistently reporting between 20,000 and 25,000 incidents annually. Traffic-related criminal code violations showed a decline from 2018, especially during 2020, where the number of reported incidents dropped below 15,000. Other categories, including "Controlled Drugs and Substances Act" and "Other Federal Statute Violations", remained consistently low over the years, each with less than 5,000 incidents reported annually.

## Crime Types by Division
Different divisions across Toronto report varying types of crimes. Divisions located in more urbanized areas, show higher concentrations of property crimes, such as auto theft and break-ins. At the same time, suburban areas report a higher frequency of violent crimes.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

crime_by_division_plot <- analysis_data %>%
  group_by(division_name, crime_subtype) %>%
  summarise(total_incidents = sum(incidents_reported, na.rm = TRUE)) # Summing incidents per division and crime type
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-Crimes_by_Div
#| fig-cap: Crimes by Division and Crime Type
#| fig-width: 10
#| fig-height: 8

## Create a bar chart comparing crime types and incidents by division
## Citation: https://allisonhorst.github.io/palmerpenguins/
## Citation: https://ggplot2.tidyverse.org
crime_by_division_plot %>%
  ggplot(aes(y = reorder(division_name, -total_incidents), x = total_incidents, fill = crime_subtype)) +
  geom_bar(stat = "identity", position = "dodge") + # Using a bar chart for comparison
  labs(
    title = "Crime Incidents by Division and Crime Type in Toronto",
    y = "Division",
    x = "Total Incidents",
    fill = "Crime Type"
  ) +
  theme_minimal() + # Minimal theme for clean visuals
  theme(
    axis.text.y = element_text(size = 8, margin = margin(r = 10)), # Add the spacing of y-axis
    plot.title = element_text(hjust = 0.5), # Placing title on the middle
    legend.position = "bottom" # Placing legend on the bottom
  ) +
  guides(fill = guide_legend(ncol = 2)) # Making the legend 2 columns
```
Based on @fig-Crimes_by_Div, 32 Division reports high rates of auto theft than other divisions. 12 Division and 53 Division show large number reports in break-ins and residential thefts, especially in apartment.

## Clearance Rates
Clearance rates, which reflect the number of resolved cases, vary widely between divisions. Many divisions report low clearance rates for property crimes, particularly auto theft, which remains a challenge for law enforcement.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

## Calculate clearance rate for each crime type and division
clearance_rate_plot <- analysis_data %>%
  filter(incidents_reported > 0) %>% # Ensure calculating clearance rate where incidents were reported
  mutate(clearance_rate = incidents_cleared / incidents_reported) %>% # Calculate the clearance rate
  group_by(crime_subtype, division_name) %>%
  summarise(avg_clearance_rate = mean(clearance_rate, na.rm = TRUE)) # Average clearance rate for each crime type and division
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-Clearance_Rate
#| fig-cap: Clearance Rate by Crime Type
#| fig-width: 10
#| fig-height: 8

## Create a grouped bar chart to show clearance rate by crime type and division
## Citation: https://allisonhorst.github.io/palmerpenguins/
## Citation: https://ggplot2.tidyverse.org
clearance_rate_plot %>%
  ggplot(aes(y = reorder(division_name, -avg_clearance_rate), x = avg_clearance_rate, fill = crime_subtype)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) + # Grouped bar chart for clearance rates
  labs(
    title = "Clearance Rate by Crime Type and Division in Toronto",
    y = "Division",
    x = "Average Clearance Rate",
    fill = "Crime Type"
  ) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) + # Convert y-axis to percentage
  theme_minimal() + # Minimal theme for clean visuals
  theme(
    axis.text.y = element_text(size = 12, margin = margin(r = 10)), # Add the spacing of y-axis
    plot.title = element_text(hjust = 0.5), # Placing title on the middle
    legend.position = "bottom" # Placing legend on the bottom
  ) +
  guides(fill = guide_legend(ncol = 2)) # Making the legend 2 columns
```
As illustrated by @fig-Clearance_Rate, the overall clearance rate for property crimes remains low, with less than 10% of auto theft cases being resolved in most divisions. Crimes against persons, such as assault and sexual violations, generally have relatively high clearance rates, particularly in suburban areas.

# Discussion {#sec-discussion}
The results of this analysis show clear trends in crime patterns across Toronto. Downtown areas  are significantly affected by property crimes like auto theft and break-ins, driven by dense population and high economic activity. Suburban areas, however, report more violent crimes, which may be influenced by socio-economic factors such as income inequality and access to social services [@Mohammadi_Bergquist].

The data also highlights the gap between reported incidents and the number of cleared cases. Low clearance rates for property crimes, particularly auto theft, suggest the need for more effective crime prevention measures. Increasing surveillance and expanding the capacity of law enforcement agencies in high-crime areas may help address this issue.

Further research into socio-economic factors driving these trends is necessary. For example, regions with higher rates of poverty and unemployment may experience more violent crimes. Additionally, strategies like community policing and public engagement in high-risk areas may prove effective in reducing crime rates and improving clearance rates.

# Conclusion {#sec-conclusion}
This paper analyzed crime trends in Toronto using the Police Annual Statistical Report from 2014 to 2023. The findings demonstrate significant varieties in crime distribution across Toronto’s police divisions, with property crimes being more concentrated in downtown areas and violent crimes more prevalent in suburban regions. These insights are critical for shaping future crime prevention policies and ensuring the effective allocation of related resources.

Further research should explore the socio-economic factors influencing these trends, including whether community engagement can help lower crime rates in high-risk areas.

\newpage

# References


